Sulfur Dioxide on Water's Surface

### 2011-06-06 EquilibratedOrientation.tex
reworked the results section. One overall results section with 3 different subsections.
Rewrote some of draft 4, but still need to finish up results and conclusions section.

\section{Equilibrated MD}

Geometric analyses were performed to characterize the net molecular orientation of \wat~and \suldiox~molecules at different depths from the water surface. At each distance from the surface location, an orientation profile was created for both the \wat~and \suldiox~molecules. The orientation distribution for the angles $\theta$ and $\phi$ at all depths were combined to form 2D intensity plots that show how the molecular orientation distributions changed with distance to the surface location. These plots allow for a visual interpretation of how the net orientations are affected when moving from the gas phase to the surface, and then further into the aqueous bulk.

\subsection{\wat~Orientation}

@@

\subsection{\suldiox~Orientation}

>>>>

\subsection{Equilibrated MD}

Geometric analyses were performed to characterize the net molecular orientation of \wat~and \suldiox~molecules at different depths from the water surface. At each distance from the surface location, an orientation profile was created for both the \wat~and \suldiox~molecules. The orientation distribution for the angles $\theta$ and $\phi$ at all depths were combined to form 2D intensity plots show how the molecular orientation distributions change with distance to the surface location. These plots allow for a visual interpretation of how the net orientations are affected when moving from the gas phase to the surface, and then further into the aqueous bulk. Both the neat-water, with only a single \suldiox~introduced, and the high-concentration saturated system were analyzed. In the case of the neat-water system, the introduction of a single \suldiox~does not greatly affect the water surface, and the results of the water orientation are very similar to a neat-water system without any adsorbed solutes.

\subsubsection{\wat~Orientation}

@@

\subsubsection{\suldiox~Orientation}

### 2011-06-06 Introduction.tex
only removed the first sentence of the intro
Rewrote some of draft 4, but still need to finish up results and conclusions section.

The physics of gaseous adsorption on water surfaces is a mostly unexplored field. Our current molecular-level understanding of adsorption phenomena relies on experiments and theory of model systems, and on an increasing number of computational simulations of molecular processes. What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps? To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process? Are specific gas or liquid molecular orientations necessary for gaseous adsorption? Very little work has been done to characterize the molecular processes that take place during the transit of a molecule from the gas phase onto a liquid interfacial region. Furthermore, experiment does not yet provide us with the resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site. In this work we look at the orientational properties of sulfur dioxide gas adsorbing to a water surface as a model for the transit of gases on to liquid interfaces.

The interactions between sulfur species and aqueous systems are currently an active area of chemical research. Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well. We seek a fuller understanding of the interactions between \suldiox~and water at interfaces because they are central to many atmospheric particle and aerosol reactions.

@@

Our latest vsfs experiments on \suldiox~behavior further expand our understanding of \suldiox~in atmospheric processes.\cite{Ota2011} A temperature study showed that the binding of gaseous \suldiox~to water surfaces is greatly enhanced at atmospherically relevant cold temperatures. The surface binding of unreacted \suldiox~complexes was found to be a completely reversible process. The same study also showed that low pH aqueous environments inhibit the bulk reaction of \suldiox, but do not affect the surface binding of the gas molecules. The resultant experimental spectra suggest that the surface waters reorient in response to the presence of adsorbed \suldiox. Knowing that reorientation may occur due to \suldiox~adsorption, how can we confirm it? what is the nature of the reorienting waters? How do they reorient, and to what extent? Conclusive experimental evidence of the adsorbing gas or surface water molecular orientation is still missing, so we are moved to use computational methods for further understanding this phenomena.

In this study we look at computed orientational depth profiles of water and \suldiox~before and after the adsorption process. Spectroscopic experiments can not probe depth or orientation profiles of surface species in the same detailed manner as computational simulations. We determine the effect of introducing \suldiox~to a \wat~surface, and analyze the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations. We use a technique of steering a gas molecule into the aqueous phase, and characterize the transit through the interfacial region. This unique method gives us much greater insight into the behaviors of gas molecules as they move near to liquid water. We also simulate the behavior of adsorbed \suldiox~at equilibrium to show how an already-adsorbed \suldiox~behaves on a water surface. The results of this study will help us to understand the behavior of adsorbing gases. Computational simulations like this are necessary to further understand the microscopic nature of the gaseous adsorption process, as many of the molecular-level subtleties are lost in experiment. Along with other ongoing experiments and computations, we develop a more complete picture of gaseous adsorption on aqueous surfaces.

>>>>

Very little work has been done to characterize the molecular processes that take place during the transit of a molecule from the gas phase onto a liquid interfacial region, and how such a surface process is related to uptake.  What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process?  Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site.
This type of information can be determined computationally, and when coupled to the experimental studies can provide important new insight into the gas-liquid suface adsorption process.

In this work we look at the orientational properties of sulfur dioxide gas adsorbing to a water surface as a model for the transit of gases on to liquid interfaces.  The interactions between sulfur species and aqueous systems are currently an active area of chemical research. Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well. We seek a fuller understanding of the interactions between \suldiox~and water at interfaces because they are central to many atmospheric particle and aerosol reactions.

@@

Our latest vsfs experiments on \suldiox~behavior further expand our understanding of \suldiox~in atmospheric processes.\cite{Ota2011} A temperature study showed that the binding of gaseous \suldiox~to water surfaces is greatly enhanced at atmospherically relevant cold temperatures. The surface binding of unreacted \suldiox~complexes was found to be a completely reversible process. The same study also showed that low pH aqueous environments inhibit the bulk reaction of \suldiox, but do not affect the surface binding of the gas molecules. The resultant experimental spectra suggest that the surface waters reorient in response to the presence of adsorbed \suldiox, but the details about this reorientation can not be garnered from the experiments.  Such details include knowing how surface water orientation changes upon approach of the gaseous \suldiox, how this varies upon \suldiox~surface bonding, and the orientation of \suldiox~throughout the surface bonding process. 
%Knowing that reorientation may occur due to \suldiox~adsorption, how can we confirm it? what is the nature of the reorienting waters? How do they reorient, and to what extent? Conclusive experimental evidence of the adsorbing gas or surface water molecular orientation is still missing, so we are moved to use computational methods for further understanding this phenomena.

In this study we look at computed orientational depth profiles of water and \suldiox~before and after the adsorption process. The studies augment the spectroscopic experiments that can not probe depth or orientation profiles of surface species in the same detailed manner as computational simulations. We determine the molecular effects that occur upon introduction of \suldiox~to a \wat~surface, and analyze the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations. With this approach we steer a gas molecule into the aqueous phase, and characterize its molecular orientation as it transits through the interfacial region.  This unique approach provides insight into the behaviors of gas molecules as they move near to liquid water. We also simulate \suldiox~at equilibrium to show how an already-adsorbed \suldiox~behaves on a water surface. The results of this study will help us to understand the behavior of adsorbing gases, allowing a more complete picture of gaseous adsorption on aqueous surfaces.

### 2011-06-06 Methods.tex
Rewrote some of draft 4, but still need to finish up results and conclusions section.

Equilibrium simulations involved adding \suldiox~to a water slab and equilibrating as outlined above. The neat-water system had a single \suldiox~added to the center of the water box, representing a concentration of 0.06 M. The saturated system with 22 \suldiox~molecules in the bulk represents a system of concentration 1.35 M, with an additional 50 \suldiox~in the gas phase to model 50 atm of pressure above the water surface. Both the low and high concentration systems were then evolved for a further 10 ns using a timestep of 0.5fs, with atomic coordinates written every 100 fs.

@@

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the equilibrated neat-water and saturated systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water 15\angs~under the water slab surface by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfuly model chemical events.\cite{Isralewitz2001} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a timestep of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

A separate set of SMD simulations were performed with tethering to one of the \suldiox-oxygens. This was done to ensure that the orientation of the \suldiox~during the adsorption transit was not an artifact of the choice of atom used for tethering. The simulations produced the same results (not shown) for the orientational analyses, so the data from the original tethering scheme were used.

@@

	In this study we create orientational depth-profiles of \suldiox~and \wat. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a lineshape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given timestep. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes. In this work we define the surface location by calculating a reference location at each timestep by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

        The location of the water surface is determined by the averaged positions of a set number of the outer-most waters in the simulated aqueous slabs. The number of waters included in the averaging is determined by taking a few issues into account. First, a visual inspection of the top cross-section of the water slab over several timesteps indicated between 65-75 waters established a full monolayer. Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres would yield a similar number of molecules to the visual inspection. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the authors decided to use 70 water molecules for calculating instantaneous water surface locations.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. This figure shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work. The water density profile and location (dashed line) is produced by averaging the instantaneous density profile at each timestep in the simulation. The water profile was then fit to a $\tanh$ function (black) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a ``90-10'' width of 8.28 $\AA$. A bulge in the gas-phase (left) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point. All depth locations in our analyses are calculated relative to the surface location (blue) at the corresponding timestep of the trajectories.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(dashed) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit (black) using a hyperbolic tangent function to extract location and width data for the water surface. (B) The location of the outer \wat~monolayer was calculated for each simulation timestep (blue), and the surface location from the density fitting was also determined (dashed). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}

The new procedure described above for finding the outer monolayer location was implemented, and the distribution of that location is also plotted in figure \ref{fig:density-flaw} (blue). It is apparent from the surface location distribution that the water slab drifts over the 10 ns, spanning approx. 12 $\AA$. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each timestep is only 1.85 $\AA$.

>>>>

Equilibrium simulations involved adding \suldiox~to a water slab and equilibrating as outlined above. The neat-water system had a single \suldiox~added to the center of the water box, representing a concentration of 0.06 M. The second, more concentrated system examined consisted of 22 \suldiox~molecules in the bulk corresponding to a concentration 1.35 M. This saturated system was exposed to an additional 50 \suldiox~in the gas phase above the water surface. Both the low and high concentration systems were then evolved for a further 10 ns using a timestep of 0.5fs, with atomic coordinates written every 100 fs.

@@

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the equilibrated neat-water and saturated systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfuly model chemical events.\cite{Isralewitz2001} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a timestep of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

A separate set of SMD simulations were performed with tethering of one of the \suldiox-oxygens to the water slab center of mass. This was done to ensure that the orientation of the \suldiox~during the adsorption transit was not an artifact of the choice of atom used for tethering. The simulations produced the same results (not shown) for the orientational analyses, so the data from the original tethering scheme were used.

@@

	In this portion of the study we create orientational depth-profiles of \suldiox~and \wat~upon exposure and adsorption of the gas. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a lineshape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given timestep. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes. In this work we define the surface location by calculating a reference location at each timestep by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

The location of the water surface is determined by the averaged positions of a set number of the outer-most waters in the simulated aqueous slabs. The number of waters included in the averaging is determined by taking a few issues into account.  First, counting the waters found in the topmost cross-section of the water slab over several timesteps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the authors decided to use 70 water molecules for calculating instantaneous water surface locations.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}a shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system. The water density profile and location (dashed line in figure \ref{fig:density-flaw}a) is produced by averaging the instantaneous density profile at each timestep in the simulation over all the timesteps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52\AA$) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

The new procedure described previously for finding the outer monolayer location was implemented, and the surface location is plotted as a function of simulation time in figure \ref{fig:density-flaw}b. It is apparent from the surface location plot that the water slab location, and thus the surface location, drifts over the 10 ns, spanning approx. 12 $\AA$. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each timestep is only 1.85 $\AA$. All depth locations in our analyses are calculated relative to the instantaneous surface location at the corresponding timestep of the trajectories.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(dashed) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit using a hyperbolic tangent function (black) to extract location and width data for the water surface. (B) The instantaneous location of the outer \wat~monolayer was calculated for each simulation timestep (blue), and the surface location extracted from the density fitting was also shown for reference (horizontal dashed line). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}
